---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
*This is the 2nd week project for Reproducible Research*

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Load required libraries

```{r load-packages}
library(tidyverse)
library(lubridate)
library(fs)
library(gt)
```


### Download the raw data

```{r load-data}
# Declare our variables
data.source <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
archive.name <- "repdata_data_activity.zip"
file.name <- "activity.csv"

# Get paths to zip file and csv file if available
zip.paths <- dir_ls(recurse = TRUE,
                   type = "file",
                   glob = paste0("*",archive.name))
file.paths <- dir_ls(recurse = TRUE,
                    type = "file",
                    glob = paste0("*",file.name)) 
                      
# download, unzip and read in file where valid                      
if (!length(zip.paths) && !length(file.paths)){
  download.file(data.source, archive.name, method="curl")
  unzip(archive.name)
  activity.raw <- read_csv(file = file.name)
  } else if (!length(file.paths)) {
    unzip(archive.name)
    activity.raw <- read_csv(file = file.name)
    } else if(!exists("activity.raw")) {
      activity.raw <- read_csv(file = file.name)
      } else {}

```

### Process the raw data

```{r process-data}
# Change date column to a date where valid.
activity.raw$date <- ymd(activity.raw$date)

```


## What is the mean total number of steps taken per day?

1. **The total number of steps taken per day for each day.**

```{r change-date}
# Group data by date and sum the steps by date
tot.steps <- activity.raw %>%
  group_by(date) %>%
  summarise(tot.steps.day = sum(steps, na.rm = TRUE))
```

2. **The Histogram of the total number of steps taken each day.**

```{r histogram}
tot.steps %>%
  ggplot(aes(x = tot.steps.day)) +
  geom_histogram(binwidth = 1000,
                 boundary = 0,
                 colour = "black",
                 fill = "grey") +
  scale_x_continuous(labels = scales::label_comma(suffix = "\nsteps")) +
  labs(y = "Count of total steps per day",
       x = "Brackets for total steps per day",
       title = "Histogram of total steps per day")
```

3. **The mean and median of the total number of steps taken per day**

```{r mean-median}
# Calculate the mean
tot.steps.mean <- tot.steps %>%
  summarise( mean.val = mean(tot.steps.day))

# Calculate the median
tot.steps %>%
  summarise(median.val = median(tot.steps.day)) %>%
  bind_cols(tot.steps.mean) %>%
  # Send values to the table
  gt() %>%
  cols_label(
    median.val = md("Median"),
    mean.val = md("Mean")
  ) %>%
  fmt_number( columns = vars(median.val, mean.val),
              decimals = 0,
              use_seps = TRUE,
              sep_mark = ",")


```


## What is the average daily activity pattern?

1. **Line Plot of the time series**

```{r line-plot}
# Group by interval and calculate the average number of steps foor each interval
avg.steps <- activity.raw %>%
  group_by(interval) %>%
  summarise(avg.steps.int = mean(steps, na.rm = TRUE))

# Generate the line plot
avg.steps %>%
  ggplot( aes(x = interval, y = avg.steps.int) ) +
  geom_line() +
  geom_point(colour = "black", alpha = 0.1) +
  scale_x_continuous(labels = scales::label_comma(suffix = "\ninterval")) +
  labs(y = "Average steps per interval",
       x = "Interval",
       title = "Line Plot of average steps per interval")
```

2. **The 5-minute interval with the maximum number of steps on average across all the days.**

```{r max-step}
# Filter for the maximum average steps per interval
avg.steps %>%
  filter(avg.steps.int == max(avg.steps.int)) %>%
  # Send value to the table
  gt() %>%
  cols_label(
    avg.steps.int = md("Maximum Average Step Value "),
    interval = md("Interval with Maximum Average Step")
  ) %>%
  fmt_number( columns = vars(avg.steps.int, interval),
              decimals = 0,
              use_seps = TRUE,
              sep_mark = ",")
```


## Imputing missing values

1. **The total number of missing values in the database**

```{r count-NA}
# Count the number of NA's
activity.raw %>%
  summarise(na.count = sum(is.na(steps))) %>%
  gt() %>%
  cols_label(
    na.count = md("Count of NA's")
  ) %>%
  fmt_number( columns = vars(na.count),
              decimals = 0,
              use_seps = TRUE,
              sep_mark = ",")
```

2. **Fill in all of the missing values in the dataset.**
3. **Create a new dataset that is equal to the original dataset but with the missing data filled in**

```{r impute}
# Use tidy models to impute using k nearest neighbours to impute missing data
library(recipes)
imputed.rec <- recipe(date ~ steps + interval,
                      data = activity.raw) %>%
  step_knnimpute(steps)

imputed.activity <- prep(imputed.rec) %>%
  juice()
```

4. **Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day**

```{r hist}
# Calculate total steps each day
imp.tot.steps <- imputed.activity %>%
  group_by(date) %>%
  summarise(imp.tot.steps.day = sum(steps)) %>%
  ungroup

# Plot the histogram
imp.tot.steps %>%
  ggplot(aes(x = imp.tot.steps.day)) +
  geom_histogram(binwidth = 1000,
                 boundary = 0,
                 colour = "black",
                 fill = "grey") +
  scale_x_continuous(labels = scales::label_comma(suffix = "\nsteps")) +
  labs(y = "Count of total steps per day including imputed",
       x = "Brackets for total steps per day",
       title = "Histogram of total steps per day including imputed")


imp.tot.steps.mean <- imp.tot.steps %>%
  summarise(imp.mean.val = mean(imp.tot.steps.day))

imp.tot.steps %>%
  summarise(imp.median.val = median(imp.tot.steps.day)) %>%
  bind_cols(imp.tot.steps.mean) %>%
  gt() %>%
  cols_label(
    imp.median.val = md("Imputed Median"),
    imp.mean.val = md("Imputed Mean")
  ) %>%
  fmt_number( columns = vars(imp.median.val, imp.mean.val),
              decimals = 0,
              use_seps = TRUE,
              sep_mark = ",")
```
**Ans:** There are changes between the mean and median before and after imputation.
After imputing the mean shifted more than the median, interestingly the overall distribution shown on the histogram stayed more or less the same. 

## Are there differences in activity patterns between weekdays and weekends?

```{r line-plot-wkday-wkend}
imputed.activity$date <- ymd(imputed.activity$date)
imputed.activity <- imputed.activity %>% mutate( wkday.wkend = case_when(wday(date) == 1 ~ "weekend",
                                                     wday(date) == 7 ~ "weekend",
                                                     TRUE ~ "weekday"))
imputed.activity %>%
  group_by(wkday.wkend, interval) %>%
  summarise(avg.steps.int = mean(steps)) %>%
  ggplot( aes(x = interval, y = avg.steps.int) ) +
  geom_line() +
  geom_point(colour = "black", alpha = 0.1) +
  facet_wrap(~wkday.wkend, nrow = 2) +
  scale_x_continuous(labels = scales::label_comma(suffix = "\ninterval")) +
  labs(y = "Average steps per interval",
       x = "Interval",
       title = "Line Plot of average steps per interval")
```



